---
title: "Lab Exercises 1"
author: "Jongtaek Lee"
date: "2024-01-13"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = TRUE)
```

```{r }
library(tidyverse)
```

```{r warning=FALSE}
dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", skip = 2, col_types = "dcddd")
head(dm)
```

# Question 1

```{r }
dm_fm_ratio <- dm |>
                  mutate(fm_ratio = Female/Male) |>
                  filter(Age==10|Age==20|Age==30|Age==40) |>
                  select(Year:Age|fm_ratio)
dm_fm_ratio
```

```{r }
dm_fm_ratio |>
  ggplot(aes(x=Year ,y=fm_ratio, color=Age, linetype=Age)) +
  geom_line() + 
  scale_color_brewer(palette = "Set1") +
  theme_bw() +
  labs(title = "Female to Male mortality ratio over time, Ontario",
       y = "Female/Male mortality")
```

# Question 2

```{r }
dm |>
  select(Year:Age|Female) |>
  group_by(Year) |>
  summarize(Age[which.min(Female)])
```

# Question 3

```{r warning=FALSE}
dm |> 
  group_by(Age) |> 
  summarize(across(Female:Total, sd, na.rm = TRUE)) |>
  arrange(as.numeric(Age))
```

```{r }
dm2 <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt", skip = 2, col_types = "dcddd")
head(dm2)
```

```{r }
colnames(dm2) <- c("Year", "Age", "Female_pop", "Male_pop", "Total_pop")

dm_new <- dm |>
            left_join(dm2)
dm_new
```

```{r }
dm_avg <- dm_new |>
  group_by(Year) |>
  summarize(Female = sum(Female*Female_pop, na.rm=TRUE)/sum(Female_pop, na.rm=TRUE),
            Male = sum(Male*Male_pop, na.rm=TRUE)/sum(Male_pop, na.rm=TRUE)) |>
  pivot_longer(Female:Male, names_to = "Sex", values_to = "Average_rate")
dm_avg
```

```{r }
dm_avg |>
  ggplot(aes(x=Year, y=Average_rate, color=Sex)) +
  geom_line() +
  scale_color_brewer(palette = "Set1") +
  labs(title = "A trend of average mortality rates over time, Ontario",
       y = "Average Mortality rate") + 
  theme_bw()
```

Since 1975, the mortality rate for female started to keep increasing until 2000, however the one for male kept decreasing in the same period.

# Question 5

```{r warning=FALSE}
y <- dm |>
        select(Year:Female) |>
        filter(Year == 2000, as.numeric(Age) < 106)
y
```

```{r }
y_data <- log(y$Female)
```

```{r }
model <- lm(y_data ~ as.numeric(y$Age), data=y)
summary(model)
```
Population regression model:

$log(Female_{i}) = \beta_{0i} + \beta_{1i}*Age_{i} + \epsilon_{i}$

Fitted regression model:

$log(Female_{i}) = -10.062281 + 0.086897*Age_{i}$

The expected value of log of female mortality rate increases by 0.086891 for every unit increase of Age. Therefore, for a woman who gets one year older, her expected mortality rate will be exp(0.086891) = 1.090788 times of the current rate.


